Studying measurement invariance in factor analysis models via stochastic processes

### 2011-02-10 MIpaper/par_instable.tex
revised the paper: mostly smaller modifications and some readme/fixme comments
revised the code, including the artificial example: especially the DGP can
be improved using a suitably scaled version of the violation magnitude



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@41 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\abstract{The issue of measurement invariance commonly arises in
  factor-analytic contexts, with methods for assessment
  including 
  likelihood ratio tests, Lagrange multiplier tests, and
  Wald tests.  These tests all require advance definition of the number of
  groups, group 
  membership, and offending model parameters.
In this paper, we construct tests of parameter
invariance based on stochastic processes of casewise
derivatives of the likelihood function.  These tests can be viewed as
generalizations of the Lagrange multiplier test, and they are especially
useful for: (1) isolating specific parameters affected by measurement
invariance violations, and (2) identifying subgroups of individuals
that violated measurement invariance based on a continuous auxiliary
variable.  The tests are presented and illustrated in detail, along
with simulations examining the tests' abilities in controlled conditions.}

@@

@@

The assumption that parameters are
invariant across observations is a
fundamental tenet of many statistical models.  A specific type of 
parameter invariance, measurement invariance, has implications for the
general design and use of psychometric scales.
This concept is 
particularly important because violations can render the scales
useless.  
That is, if a set of scales violates measurement invariance,
then individuals with the same ``amount'' of a latent variable 
may systematically receive 
different scale scores.
This may lead researchers to conclude subgroup
differences on a wide variety of interesting constructs 
when, in reality, the scales are the sole cause of the 
differences.  Further, it can be inappropriate to incorporate scales
violating measurement invariance into structural equation
models, where relationships between constructs are hypothesized.  Horn
and McArdle \citeyear{HorMca92} concisely summarize the impact of
these issues, stating ``Lack of evidence of measurement 
invariance equivocates conclusions and casts doubt on theory in the
behavioral sciences'' (p.\ 141).  Borsboom \citeyear{Bor06a} also
notes that researchers often fail to assess whether measurement
invariance holds.

@@

To frame this as a formal hypothesis, we assume that -- in principle --
Model~(\ref{eq:loglik}) holds for all individuals but with a
potentially individual-specific 
parameter vector ${\bm \theta}_i$. The null hypothesis of measurement invariance
is then equivalent to the null hypothesis of parameter constancy
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
which should be tested against the alternative that the parameters
are some nonconstant function ${\bm \theta}(\cdot)$ of the variable $V$ with observations
$v_1, \dots, v_n$, i.e.,
\begin{equation}
    \label{eq:h1}
    H_1:~ {\bm \theta}_i = {\bm \theta}(v_i),\quad (i=1,\ldots,n).
\end{equation}
where the pattern ${\bm \theta}(V)$ of deviation from measurement invariance is
typically not known (exactly) in practice. If it were (see below for
some concrete examples), then standard inference methods (such as likelihood
ratio, Wald, or Lagrange multiplier tests) could be employed. However,
if the pattern is unknown, it is difficult to develop a single test
that is well-suited for all conceivable patterns. But it is possible to
derive a family of tests so that representatives from this family are well-suited
for a wide range of possible patterns. One pattern of particular interest
involves $V$ dividing the individuals into two subgroups with different
parameter vectors
\begin{equation}
  \label{eq:h1*}
  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where ${\bm \theta}^{(A)} \neq {\bm \theta}^{(B)}$. This could pertain to 
two different age groups, income groups, genders, etc.

Note that even when adopting $H_1^*$ as the relevant alternative, the pattern ${\bm \theta}(V)$
is not completely specified unless the cutpoint $\nu$ is known in
advance. 
In this situation, all individuals can be grouped based on
$V$, and we can apply standard
theory: nested multiple group
models (e.g., \citeNP{Jor71,Bol89}) coupled with likelihood ratio (LR)
tests are most common,
though Lagrange multiplier (LM) and Wald tests may also be constructed for this
purpose.\footnote{All three of these approaches are asymptotically
  equivalent \cite{Sat89}.}  However, if $\nu$ is unknown (as is often
the case for continuous $V$), then the standard theory does not apply
and we can instead rely on the tests proposed in this paper.

@@

% MAY BE USEFUL SOMEWHERE
% Focusing on the latter point, Millsap's \citeyear{Mil05} review of 
% measurement invariance methods cites ``locating the violation of the
% invariance'' as one of the major outstanding problems.  The tests
% resulting from the proposed project directly address this issue,
% leading to an important advance in measurement invariance research.

@@

To employ the LR test for assessing measurement invariance, the data
are typically separated into a certain number of given subgroups where
the parameters are estimated separately (for ease of
  exposition, we assume no parameters are restricted to be equal
  across subgroups). Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.
\readme{Achim: You previously modified the above section so that $\theta$
  consists only of parameters that are ``doubled'' under the
  alternative.  This appears to not allow for situations where some 
  parameters are restricted to be equal across subgroups.  This should
  work for the current presentation, 
  but many applications of LR tests with these models have
  some parameters restricted to be equal across subgroups.  So I've
  added some detail in parentheses at the beginning of the above
  paragraph.}

@@

Instead of choosing a specific $\nu$, a natural idea is to compute $\mathit{LR}(\nu)$
for each possible value in some interval $[\underline{\nu}, \overline{\nu}]$ and
reject if their maximum
\begin{equation} \label{eq:maxlr}
  \max_{\nu \in [\underline{\nu}, \overline{\nu}]} \mathit{LR}(\nu)
\end{equation}
becomes large. Note that this corresponds to maximizing the likelihood w.r.t.\
an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. However, Andrews \citeyear{And93}
showed that the asymptotic distribution is in fact tractable, but nonstandard.
Specifically, the asymptotic distribution of the maximum is the maximum of a
certain tied-down Bessel process whose specifics also depend on the minimal
and maximal thresholds $\underline{\nu}$ and $\overline{\nu}$, respectively.
See below for more details.

@@

So far, the discussion focused on the alternative $H_1^*$: The maximum LR, Wald, and LM
tests are designed for a situation where there is a single threshold at which the full
parameter vector ${\bm \theta}$ changes. While this is plausible and intuitive
in many applications, it would also be desirable to obtain tests that direct their power
against other types of alternatives, i.e., against $H_1$ with other patterns ${\bm \theta}(V)$.
For example, there might be multiple thresholds at which the parameters change or parameters
fluctuate randomly. Alternatively, only one (or just a few) of the parameters in
the vector ${\bm \theta}$ change while the remaining parameters are
constant (a common occurrence in psychometric models).
To address such situations in a unified way, the next section contains 
a general framework for testing measurement invariance along a
(continuous) variable $V$ that
includes the maximum LM test as a special case.

@@

As argued above, no single aggregation function will have high power for any
conceivable pattern of measurement invariance ${\bm \theta}(V)$ while any
(reasonable) aggregation function will have non-trivial power under $H_1$.
Thus, various aggregation strategies should be employed depending on which
pattern ${\bm \theta}(V)$ is most plausible (because the exact is typically
unknown). A particularly agnostic aggregation strategy is to reject $H_0$
if any component of the the cumulative score process ${\bm B}(t; \hat {\bm \theta})$
departs ``too far'' from its zero at any time, i.e., if
\begin{equation}
    \label{eq:dmax}
    \mathit{DM} = \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,
\end{equation}
becomes large. Consquently, this double maximum statistic allows for simultaneous
isolation of the threshold(s) of parameter change (over the individuals $i = 1, \dots, n$)
and the parameter(s) affected by it (over $j = 1, \dots, k$). This test is
especially useful for visualization, as the cumulative score process for
each individual parameter can be displayed along with the appropriate
critical value.  An example of this visualization appears in the
Example section.

However, taking maximums ``wastes'' power if many of the $k$~parameters change
at the same threshold or if the score process takes large values for
many of the $n$~individuals (and just a single threshold). In such
cases, sums insteads of maximums are more suitable for collapsing across
parameters and/or individuals because they combine the deviations instead of
picking out only the single largest deviation. Thus, if the parameter instability
${\bm \theta}(V)$ affects many parameters and leads to many subgroups, sums
of (absolute or squared) values should be used for collapsing both across
parameters and individuals. On the other hand, if there is just a single
threshold that affects multiple parameters, then the natural aggregation
is by sums over parameters and then by the maximum over individuals.
More precisely, the former idea leads to a Cram{\'e}r-von Mises type
statistic and the latter to the maximum LM statistic from the previous section:
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
where the $\max \mathit{LM}$ statistic is additionally scaled by the
asymptotic variance $t (1 - t)$ of the process ${\bm B}(t, \hat {\bm \theta})$.
It is equivalent to the $\max_\nu \mathit{LM}(\nu)$ from the previous
section (provided that the boundaries for the subgroups sizes $\underline{i}$/$\underline{\nu}$
and $\overline{\imath}$/$\overline{\nu}$ are chosen analogously).

@@

@@

For the maximum LR/Wald/LM tests, it is natural to graph the sequence of LR/Wald/LM
statistics along $V$ along with a boundary corresponding to the critical value.
Again, a boundary crossing signals a significant violation, and the peak(s) in the
sequence of statistics conveys threshold(s). Note that due to summing over all
parameters, no specific parameter can be identified that is responsible for the
violation. Similarly, for the Cram\'{e}r-von Mises test neither component(s) nor threshold(s)
can be formally identified. However, graphing of (transformations of) the cumulative
score process may still be valuable for gaining some insights.

If a measurement invariance violation is detected by any of the tests,
one may want to incorporate it into the model to account for it. How this should
be done typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
discussed above often prove helpful in determining a suitable parametrization.
In particular, one approach that is often employed in practice is to adopt a model
with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
single threshold case). In the multiple threshold case, their location can
be determined by maximizing the corresponding segmented log-likelihood over all
possible combinations of thresholds (\citeNP{ZeiSha10} adapt a dynamic programming
algorithm to this task). For the single threshold case, this reduced to
maximizing the segmented log-likelihood
\begin{equation} \label{eq:seglik}
  \ell(\hat {\bm \theta}^{({A)}}; x_1, \dots, x_m) ~+~ \ell(\hat {\bm \theta}^{({B)}}; x_{m+1}, \dots, x_n)
\end{equation}
over all values of $m$ corresponding to possible thresholds $\nu$ (such that $v_m \le \nu$ and
$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR statistic
from (\ref{eq:lr}) and some minimal subgroup size is typically imposed.

@@

\begin{figure}
\caption{Path diagram representing the base factor analysis model used for
  the Example and Simulations.  To induce measurement invariance
  violations, a seventh observed variable (student age) determines the
values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
$\lambda_{31}$).}
\label{fig:famod}
\includegraphics[height=5in]{famod.pdf}
%\input{figs/path}
\end{figure}

@@

@@

To formally represent these ideas, we specify that the data arise from
a factor analysis model with two factors.  The base model, displayed in
Figure~\ref{fig:famod}, specifies that measurement invariance holds,
with three scales arising from the verbal factor and three 
scales arising from the mathematical factor.
For measurement invariance violations, we specify that $V$
(student age) impacts the values of 
verbal factor loadings in the model: 
if students are 16 through 18 years of age, then the factor
loadings corresponding to the first 
factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
Figure~\ref{fig:famod}.  If students are 13 through 15 years of age,
however, then the factor loadings corresponding to the first factor
are three standard errors lower than those in Figure~\ref{fig:famod}
(the standard errors were obtained by fitting
  the model to a sample of size 200, where the sample consisted of all
  ``older'' subjects).
This violation states that the verbal ability scales
lack measurement invariance with respect to age.  For simplicity, we
assume that the mathematical scales are invariant.

@@

Note that the tests give us the flexibility to test various subsets of
parameters.  For example, were we only interested in the 
three verbal factor loading parameters, we can test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{11}\ \lambda_{21}\ \lambda_{31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$.  Alternatively, we
can consider all model parameters (including means), leading to a test
of \eqref{eq:h0}.  We consider both of these tests below.

\subsection{Results}
In the results section, we first describe overall results.  We then 
describe estimation of $\nu$, and 
isolation of model parameters violating
measurement invariance.  R code for obtaining these results is
  available at \url{http://semtools.r-forge.r-project.org}.

@@

The figures display many test attributes, all of which generally
conform with the underlying theory.  First, when we test
only the three verbal factor loadings, all three tests can detect the
change.  Conversely, when all nineteen model parameters are tested,
only the double-max test detects a change.  This implies that 
power increases as larger proportions of tested parameters
change.  The results also imply that the double-max test is more
sensitive to fluctuations among a small subset of parameters.

Figure~\ref{fig:lrlm} 
compares the max-LM statistic (black line) to the max-LR
statistic (blue line) from \eqref{eq:maxlr}, as applied to testing
\eqref{eq:h0}.  The critical values for these two
tests are identical, hence the single horizontal line.
The figure
shows that the two statistics are very similar to one another, with
both maxima at the dotted vertical line.  This
is generally to be expected, because the two tests are asymptotically
equivalent.  For this specific situation, however, the max-LR
statistic rejects \eqref{eq:h0} while the max-LM statistic does not.
More generally, the 
max-LR statistic cannot be obtained from the empirical fluctuation
process, so it must be computed ``by hand'' (i.e., by computing
the LR statistic over all values of the threshold $\nu$, re-fitting the
factor analysis model for each value).

\subsubsection{Estimation of $\nu$ and Parameter Isolation}
As described above, the tests of \eqref{eq:exh0} imply that the 
verbal scales lack measurement invariance.  
We can also use the tests to: (1) estimate the 
threshold $\nu$, and (2) isolate specific parameters that violate
measurement invariance.  
For example, as described previously, estimates of $\nu$ can be obtained by
examining the peaks in Figure~\ref{fig:cusumex}.  For all six panels
in the figure, the peaks occur near an age of 16.1.  This agrees well
with the true threshold of 16.0.

@@

\begin{figure}
\caption{A comparison of the max-LM (black line) and max-LR (blue line) test
  statistics for \eqref{eq:h0}.  The solid, horizontal line
  corresponds to the critical value at $\alpha=.05$.}
\label{fig:lrlm}
\includegraphics[height=5in]{lmlr_19.pdf}
\end{figure}

>>>>

\abstract{
  The issue of measurement invariance commonly arises in factor-analytic
  contexts, with methods for assessment including likelihood ratio tests,
  Lagrange multiplier tests, and Wald tests. These tests all require advance
  definition of the number of groups, group membership, and offending model
  parameters. In this paper, we construct tests of parameter invariance based on
  stochastic processes of casewise derivatives of the likelihood function. These
  tests can be viewed as generalizations of the Lagrange multiplier test, and they
  are especially useful for: (1)~isolating specific parameters affected by
  measurement invariance violations, and (2)~identifying subgroups of individuals
  that violated measurement invariance based on a continuous auxiliary variable. 
  The tests are presented and illustrated in detail, along with simulations
  examining the tests' abilities in controlled conditions.
}

@@

\spacing{1}

@@

The assumption that parameters are
invariant across observations is a
fundamental tenet of many statistical models.  A specific type of 
parameter invariance, measurement invariance, has implications for the
general design and use of psychometric scales.
This concept is 
particularly important because violations can render the scales
useless.  
That is, if a set of scales violates measurement invariance,
then individuals with the same ``amount'' of a latent variable 
may systematically receive 
different scale scores.
This may lead researchers to conclude subgroup
differences on a wide variety of interesting constructs 
when, in reality, the scales are the sole cause of the 
differences.  Further, it can be inappropriate to incorporate scales
violating measurement invariance into structural equation
models, where relationships between constructs are hypothesized.  Horn
and McArdle \citeyear{HorMca92} concisely summarize the impact of
these issues, stating ``Lack of evidence of measurement 
invariance equivocates conclusions and casts doubt on theory in the
behavioral sciences'' (p.~141).  Borsboom \citeyear{Bor06a} also
notes that researchers often fail to assess whether measurement
invariance holds.

@@

To frame this as a formal hypothesis, we assume that -- in principle --
Model~(\ref{eq:loglik}) holds for all individuals but with a
potentially individual-specific 
parameter vector ${\bm \theta}_i$. The null hypothesis of measurement invariance
is then equivalent to the null hypothesis of parameter constancy
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
which should be tested against the alternative that the parameters
are some nonconstant function ${\bm \theta}(\cdot)$ of the variable $V$ with observations
$v_1, \dots, v_n$, i.e.,
\begin{equation}
    \label{eq:h1}
    H_1:~ {\bm \theta}_i = {\bm \theta}(v_i),\quad (i=1,\ldots,n).
\end{equation}
where the pattern ${\bm \theta}(V)$ of deviation from measurement invariance is
typically not known (exactly) in practice. If it were (see below for
some concrete examples), then standard inference methods -- such as likelihood
ratio, Wald, or Lagrange multiplier tests -- could be employed. However,
if the pattern is unknown, it is difficult to develop a single test
that is well-suited for all conceivable patterns. But it is possible to
derive a family of tests so that representatives from this family are well-suited
for a wide range of possible patterns. One pattern of particular interest
involves $V$ dividing the individuals into two subgroups with different
parameter vectors
\begin{equation}
  \label{eq:h1*}
  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where ${\bm \theta}^{(A)} \neq {\bm \theta}^{(B)}$. This could pertain to 
two different age groups, income groups, genders, etc.

Note that even when adopting $H_1^*$ as the relevant alternative, the pattern
${\bm \theta}(V)$ is not completely specified unless the cutpoint $\nu$ is known
in advance.  In this situation, all individuals can be grouped based on $V$, and
we can apply standard theory: nested multiple group models (e.g.,
\citeNP{Jor71,Bol89}) coupled with likelihood ratio (LR) tests are most common,
although the asymptotically equivalent (see \citeNP{Sat89}) Lagrange multiplier
(LM) and Wald tests may also be constructed for this purpose. However, if $\nu$
is unknown (as is often the case for continuous $V$), then the standard theory
cannot be easily applied and instead nonstandard inference methods are required
such as the tests proposed in this paper. \readme{Slightly rephrase but probably
still not optimal.}

@@

@@

To employ the LR test for assessing measurement invariance, the data
are typically separated into a certain number of given subgroups where
the parameters are estimated separately, potentially with some parameters
restricted to be equal across subgroups. For ease of exposition we present
the results for the case where there are no such restrictions (although it is
straightforward to extend all methods to the more general case).
Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.

@@

Instead of choosing a specific $\nu$, a natural idea is to compute $\mathit{LR}(\nu)$
for each possible value in some interval $[\underline{\nu}, \overline{\nu}]$ and
reject if their maximum
\begin{equation} \label{eq:maxlr}
  \max_{\nu \in [\underline{\nu}, \overline{\nu}]} \mathit{LR}(\nu)
\end{equation}
becomes large. Note that this corresponds to maximizing the likelihood w.r.t.\
an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. However, Andrews \citeyear{And93}
showed that the asymptotic distribution is in fact tractable, but nonstandard.
Specifically, the asymptotic distribution of (\ref{eq:maxlr}) is the maximum of a
certain tied-down Bessel process whose specifics also depend on the minimal
and maximal thresholds $\underline{\nu}$ and $\overline{\nu}$, respectively.
See below for more details.

@@

So far, the discussion focused on the alternative $H_1^*$: The maximum LR, Wald, and LM
tests are designed for a situation where there is a single threshold at which all
parameters in the vector ${\bm \theta}$ change. While this is plausible and intuitive
in many applications, it would also be desirable to obtain tests that direct their power
against other types of alternatives, i.e., against $H_1$ with other patterns ${\bm \theta}(V)$.
For example, the parameters may fluctuate randomly or there might be multiple thresholds
at which the parameters change. Alternatively, only one (or just a few) of the parameters in
the vector ${\bm \theta}$ change while the remaining parameters are
constant (a common occurrence in psychometric models).
To address such situations in a unified way, the next section contains 
a general framework for testing measurement invariance along a
(continuous) variable $V$ that
includes the maximum LM test as a special case.

@@

As argued above, no single aggregation function will have high power for any
conceivable pattern of measurement invariance ${\bm \theta}(V)$ while any
(reasonable) aggregation function will have non-trivial power under $H_1$.
Thus, various aggregation strategies should be employed depending on which
pattern ${\bm \theta}(V)$ is most plausible (because the exact is typically
unknown). A particularly agnostic aggregation strategy is to reject $H_0$
if any component of the the cumulative score process ${\bm B}(t; \hat {\bm \theta})$
departs ``too far'' from its zero at any time, i.e., if
\begin{equation}
    \label{eq:dmax}
    \mathit{DM} = \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,
\end{equation}
becomes large. Consquently, this double maximum statistic allows for simultaneous
isolation of the threshold(s) of parameter change (over the individuals $i = 1, \dots, n$)
and the parameter(s) affected by it (over $j = 1, \dots, k$). This test is
especially useful for visualization, as the cumulative score process for
each individual parameter can be displayed along with the appropriate
critical value.  An example of this visualization appears in the
example section. \readme{Maybe give the explicit figure reference?}

However, taking maximums ``wastes'' power if many of the $k$~parameters change
at the same threshold or if the score process takes large values for
many of the $n$~individuals (and not just a single threshold). In such
cases, sums insteads of maximums are more suitable for collapsing across
parameters and/or individuals because they combine the deviations instead of
picking out only the single largest deviation. Thus, if the parameter instability
${\bm \theta}(V)$ affects many parameters and leads to many subgroups, sums
of (absolute or squared) values should be used for collapsing both across
parameters and individuals. On the other hand, if there is just a single
threshold that affects multiple parameters, then the natural aggregation
is by sums over parameters and then by the maximum over individuals.
More precisely, the former idea leads to a Cram{\'e}r-von Mises type
statistic and the latter to the maximum LM statistic from the previous section:
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
where the $\max \mathit{LM}$ statistic is additionally scaled by the
asymptotic variance $t (1 - t)$ of the process ${\bm B}(t, \hat {\bm \theta})$.
It is equivalent to the $\max_\nu \mathit{LM}(\nu)$ from the previous
section (provided that the boundaries for the subgroups sizes $\underline{i}$/$\underline{\nu}$
and $\overline{\imath}$/$\overline{\nu}$ are chosen analogously).

@@

Finally, all tests can be easily modified to address the situation of
so-called ``partial structural changes'' \cite{And93}. This refers to the case of some
parameters being known to be stable, i.e., restricted to be constant across
potential subgroups. Tests for potential changes/instabilities only in the
remaining parameters are easily constructed by omitting those columns from
${\bm B}(\hat {\bm \theta})_{ij}$ that are restricted/stable and only retaining those
that are potentially instable. \readme{I guess we should say something like
this in the theory section. Is this intelligible enough?}

@@

For the maximum LR/Wald/LM tests, it is natural to graph the sequence of LR/Wald/LM
statistics along $V$ along with a boundary corresponding to the critical value.
Again, a boundary crossing signals a significant violation, and the peak(s) in the
sequence of statistics conveys threshold(s). Note that due to summing over all
parameters, no specific parameter can be identified that is responsible for the
violation. Similarly, for the Cram\'{e}r-von Mises test neither component(s) nor threshold(s)
can be formally identified. However, graphing of (transformations of) the cumulative
score process may still be valuable for gaining some insights.  \readme{Maybe give figure
references for the example?}

If a measurement invariance violation is detected by any of the tests,
one may want to incorporate it into the model to account for it. How this should
be done typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
discussed above often prove helpful in determining a suitable parametrization.
In particular, one approach that is often employed in practice is to adopt a model
with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
single threshold case). In the multiple threshold case, their location can
be determined by maximizing the corresponding segmented log-likelihood over all
possible combinations of thresholds (\citeNP{ZeiSha10} adapt a dynamic programming
algorithm to this task). For the single threshold case, this reduces to
maximizing the segmented log-likelihood
\begin{equation} \label{eq:seglik}
  \ell(\hat {\bm \theta}^{({A)}}; x_1, \dots, x_m) ~+~ \ell(\hat {\bm \theta}^{({B)}}; x_{m+1}, \dots, x_n)
\end{equation}
over all values of $m$ corresponding to possible thresholds $\nu$ (such that $v_m \le \nu$ and
$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR statistic
from (\ref{eq:lr}) and some minimal subgroup size is typically imposed.

@@

\begin{figure}
\caption{Path diagram representing the base factor analysis model used for
  the example and simulations.  To induce measurement invariance
  violations, a seventh observed variable (student age) determines the
values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
$\lambda_{31}$).}
\label{fig:famod}
\includegraphics[height=5in]{famod.pdf}
%\input{figs/path}
\end{figure}

@@

\fixme{I think it would be useful to say early on that a single static
data set is drawn from the model first for illustration -- and subsequently
a full simulation study using the same scenario is considered.}

@@

To formally represent these ideas, we specify that the data arise from
a factor analysis model with two factors.  The base model, displayed in
Figure~\ref{fig:famod}, specifies that measurement invariance holds,
with three scales arising from the verbal factor and three 
scales arising from the mathematical factor.
For measurement invariance violations, we specify that $V$
(student age) impacts the values of 
verbal factor loadings in the model: 
if students are 16 through 18 years of age, then the factor
loadings corresponding to the first 
factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
Figure~\ref{fig:famod}.  If students are 13 through 15 years of age,
however, then the factor loadings corresponding to the first factor
are three standard errors lower than those in Figure~\ref{fig:famod}
(the standard errors were obtained by fitting
  the model to a sample of size 200, where the sample consisted of all
  ``older'' subjects).
\fixme{This wasn't fully clear to me before and I probably should
have asked earlier. I think we should employ the asymptotic standard
errors divided by $\sqrt{n}$.}
This violation states that the verbal ability scales
lack measurement invariance with respect to age.  For simplicity, we
assume that the mathematical scales are invariant.

@@

Note that the tests give us the flexibility to test various subsets of
parameters.  For example, were we only interested in the 
three verbal factor loading parameters, we can test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{0, 11}\ \lambda_{0, 21}\ \lambda_{0, 31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$.  Alternatively, we
can consider all model parameters (including means), leading to a test
of \eqref{eq:h0}.  We consider both of these tests below.

\fixme{I think the paragraph above should say something like ``If we
knew or suspected which coefficients are subject to change\dots'' or
something else along those lines. Just to make clear that we assume
additional knowledge here. Also it may help to refer back to the
paragraph that I added in the theory section about partial changes.}

\subsection{Results}
In the results section, we first describe overall results.  We then 
describe estimation of $\nu$, and 
isolation of model parameters violating
measurement invariance.

@@

The figures convey information about several properties
of the tests. First, all three tests are more powerful (and in
this example significant) if only those parameters are tested
that are subject to instabilities. Conversely, if all 19~model
parameters are assessed (including those that are in fact invariant),
the power is decreased. This decrease, however, is less pronounced
for the double-max test as it is more 
sensitive to fluctuations among a small subset of parameters.

Figure~\ref{fig:lrlm} 
compares the max-LM statistic (solid line) to the max-LR
statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
\eqref{eq:h0}.  The critical values for these two
tests are identical, hence the single horizontal line.
The figure
shows that the two statistics are very similar to one another, with
both maxima at the dotted vertical line.  This
is generally to be expected, because the two tests are asymptotically
equivalent.  More generally, the 
max-LR statistic cannot be obtained from the empirical fluctuation
process, so the factor analysis model must be refitted before and after
each of the possible threshold values $\nu$.
\readme{With the modified DGP code, both max-LR and max-LM, are
nonsignificant in the example.}

\subsubsection{Estimation of $\nu$ and Parameter Isolation}
As described above, the tests of \eqref{eq:exh0} imply that the 
verbal scales lack measurement invariance.  
We can also use the tests to: (1)~estimate the 
threshold $\nu$, and (2)~isolate specific parameters that violate
measurement invariance.  
For example, as described previously, estimates of $\nu$ can be obtained by
examining the peaks in Figure~\ref{fig:cusumex}.  For all six panels
in the figure, the peaks occur near an age of 16.1.  This agrees well
with the true threshold of 16.0.

\fixme{I'm not really happy with the following paragraph. I think
I wouldn't call this a posthoc test. It is really a test of the same
hypothesis as before, but now using the double-max
statistic. Would it be possible to simply show this and say that
it is an advantage of the double-max test that both the threshold
and the parameter affected by the invariance can be ``seen'' in the
visualization. Then, we can still go on and talk about the power issue:
The double-max test potentially wastes a lot of power, namely
in the case where many of the parameters change at the same time. Note
that in the case where all parameters change, we wouldn't have to ask
the question ``Which parameter did change?''. Thus, luckily, the double-max test
is more powerful when we need it. But, unfortunately, we don't know that
in advance.}

@@

\begin{figure}
\caption{A comparison of the max-LM (solid line) and max-LR (dashed line) test
  statistics for \eqref{eq:h0}.  The solid horizontal line
  corresponds to the critical value at $\alpha=.05$ while the dotted vertical
  line highlights the threshold at which the test statistics are maximized.}
\label{fig:lrlm}
\includegraphics[height=5in]{lmlr_19.pdf}
\end{figure}
